---
title: "TÜİK Foreign Trade Indices Forecasting Project"
author: "Zeynep Sena Heptaş"
date: "2026-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Project Overview

This project forecasts Turkey's Export Unit Value Index [2015=100] using monthly
data obtained from the TÜİK Data Portal. Data access is performed reproducibly
through the `tuikr` R package (theme 17: International Trade).

# 2. Data Access via tuikr

```{r data-access}
library(tuikr)
library(dplyr)
library(readxl)
library(httr)

# Use tuikr to access TÜİK International Trade theme (ID: 17)
dis_ticaret_tables <- statistical_tables("17")

# Retrieve the official TÜİK URL for Foreign Trade Indices (istab type)
data_url <- dis_ticaret_tables |>
  filter(node_type == "istab") |>
  filter(grepl("Foreign Trade Indices", table_name)) |>
  slice(1) |>
  pull(table_url)

cat("TÜİK Theme: International Trade (17)\n")
cat("Table: Foreign Trade Indices\n")
cat("URL obtained via tuikr:\n", data_url, "\n\n")
cat("Note: Direct HTTP GET from Posit Cloud returns HTTP 403 due to TÜİK IP\n")
cat("restrictions. The file was downloaded via browser using the URL above.\n")
cat("No manual editing was performed on the raw file.\n")

# Load the raw file
df_raw <- read_excel("Foreign Trade Indices.xls", sheet = 1)
cat("\nRaw data loaded:", nrow(df_raw), "rows,", ncol(df_raw), "columns\n")
```

# 3. Data Preparation

```{r data-prep}
library(tidyr)
library(lubridate)

# Rows 6-19 contain Export Unit Value Index values (years 2013-2026)
df_ihracat <- df_raw[6:19, 1:13]
colnames(df_ihracat) <- c("Year","Jan","Feb","Mar","Apr","May","Jun",
                           "Jul","Aug","Sep","Oct","Nov","Dec")

# Convert to long (tidy) format
df_long <- df_ihracat |>
  pivot_longer(cols = Jan:Dec, names_to = "Month", values_to = "Value") |>
  mutate(
    Value     = as.numeric(Value),
    Month_num = match(Month, c("Jan","Feb","Mar","Apr","May","Jun",
                               "Jul","Aug","Sep","Oct","Nov","Dec")),
    Date      = as.Date(paste(Year, Month_num, "01", sep = "-"))
  ) |>
  filter(!is.na(Value)) |>
  arrange(Date)

# Create ts object (monthly, starting January 2013)
ts_data <- ts(df_long$Value, start = c(2013, 1), frequency = 12)
n_obs   <- length(ts_data)
ts_vec  <- as.numeric(ts_data)

cat("=== DATA SUMMARY ===\n")
cat("Variable        : Export Unit Value Index [2015=100]\n")
cat("TÜİK Theme      : International Trade (Theme 17)\n")
cat("Table           : Foreign Trade Indices\n")
cat("Frequency       : Monthly\n")
cat("Coverage        :", as.character(min(df_long$Date)),
    "to", as.character(max(df_long$Date)), "\n")
cat("Latest obs.     :", as.character(max(df_long$Date)), "\n")
cat("Forecast target : 2026-04-01 (April 2026)\n")
cat("Observations    :", n_obs, "\n")
cat("Data access date: 2026-05-23\n")
```

# 4. Exploratory Analysis

```{r eda}
library(ggplot2)
dir.create("outputs/figures", recursive = TRUE, showWarnings = FALSE)
dir.create("outputs/tables",  recursive = TRUE, showWarnings = FALSE)

# Actual series plot
p_actual <- ggplot(df_long, aes(x = Date, y = Value)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  labs(title    = "Export Unit Value Index [2015=100] — Monthly",
       subtitle = "TÜİK International Trade Data, 2013–2026",
       x = "Date", y = "Index Value") +
  theme_minimal()
print(p_actual)
ggsave("outputs/figures/actual_series_plot.png", p_actual, width = 10, height = 5)

# Classical decomposition
plot(decompose(ts_data))

cat("\nMean :", round(mean(ts_vec), 2), "\n")
cat("SD   :", round(sd(ts_vec),   2), "\n")
cat("Min  :", round(min(ts_vec),  2), "\n")
cat("Max  :", round(max(ts_vec),  2), "\n")
cat("\nThe series shows an upward trend from 2020 onward and mild seasonality.\n")
```

# 5. Forecasting Methods

## 5.1 Naïve Forecasting

```{r naive}
library(forecast)

naive_fit    <- naive(ts_data, h = 1)
naive_fitted <- c(NA, head(ts_vec, -1))
naive_err    <- ts_vec - naive_fitted

naive_bias <- mean(naive_err, na.rm = TRUE)
naive_mad  <- mean(abs(naive_err), na.rm = TRUE)
naive_mse  <- mean(naive_err^2,    na.rm = TRUE)
naive_mape <- mean(abs(naive_err / ts_vec) * 100, na.rm = TRUE)
naive_rsfe <- sum(naive_err, na.rm = TRUE)
naive_ts   <- naive_rsfe / naive_mad
naive_next <- as.numeric(naive_fit$mean[1])

cat("Naive — Bias:", round(naive_bias,4), "| MAD:", round(naive_mad,4),
    "| MSE:", round(naive_mse,4), "| MAPE:", round(naive_mape,4),
    "| RSFE:", round(naive_rsfe,4), "| TS:", round(naive_ts,4),
    "| Forecast:", round(naive_next,4), "\n")

p_naive <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,        color = "Actual")) +
  geom_line(aes(y = naive_fitted, color = "Naïve"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","Naïve"="red")) +
  labs(title = "Actual vs Naïve Forecast", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_naive)
ggsave("outputs/figures/naive_forecast_plot.png", p_naive, width = 10, height = 5)
```

## 5.2 Moving Average

```{r moving-average}
ma_k      <- 3
ma_fitted <- as.numeric(stats::filter(ts_vec, rep(1/ma_k, ma_k), sides = 1))
ma_err    <- ts_vec - ma_fitted

ma_bias <- mean(ma_err, na.rm = TRUE)
ma_mad  <- mean(abs(ma_err), na.rm = TRUE)
ma_mse  <- mean(ma_err^2,    na.rm = TRUE)
ma_mape <- mean(abs(ma_err / ts_vec) * 100, na.rm = TRUE)
ma_rsfe <- sum(ma_err, na.rm = TRUE)
ma_ts   <- ma_rsfe / ma_mad
ma_next <- mean(tail(ts_vec, ma_k))

cat("MA(3) — Bias:", round(ma_bias,4), "| MAD:", round(ma_mad,4),
    "| MSE:", round(ma_mse,4), "| MAPE:", round(ma_mape,4),
    "| RSFE:", round(ma_rsfe,4), "| TS:", round(ma_ts,4),
    "| Forecast:", round(ma_next,4), "\n")

p_ma <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,     color = "Actual")) +
  geom_line(aes(y = ma_fitted, color = "MA(3)"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","MA(3)"="darkorange")) +
  labs(title = "Actual vs Moving Average (k=3)", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_ma)
ggsave("outputs/figures/moving_average_plot.png", p_ma, width = 10, height = 5)
```

## 5.3 Weighted Moving Average

```{r wma}
# Weights: 1/6, 2/6, 3/6 — more weight on recent observations
w         <- c(1/6, 2/6, 3/6)
wma_fitted <- rep(NA_real_, n_obs)
for (i in 3:n_obs) wma_fitted[i] <- sum(w * ts_vec[(i-2):i])
wma_err  <- ts_vec - wma_fitted

wma_bias <- mean(wma_err, na.rm = TRUE)
wma_mad  <- mean(abs(wma_err), na.rm = TRUE)
wma_mse  <- mean(wma_err^2,    na.rm = TRUE)
wma_mape <- mean(abs(wma_err / ts_vec) * 100, na.rm = TRUE)
wma_rsfe <- sum(wma_err, na.rm = TRUE)
wma_ts   <- wma_rsfe / wma_mad
wma_next <- sum(w * ts_vec[(n_obs-2):n_obs])

cat("WMA — Bias:", round(wma_bias,4), "| MAD:", round(wma_mad,4),
    "| MSE:", round(wma_mse,4), "| MAPE:", round(wma_mape,4),
    "| RSFE:", round(wma_rsfe,4), "| TS:", round(wma_ts,4),
    "| Forecast:", round(wma_next,4), "\n")

p_wma <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,      color = "Actual")) +
  geom_line(aes(y = wma_fitted, color = "WMA"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","WMA"="purple")) +
  labs(title = "Actual vs Weighted Moving Average", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_wma)
ggsave("outputs/figures/weighted_moving_average_plot.png", p_wma, width = 10, height = 5)
```

## 5.4 Exponential Smoothing

```{r exp-smoothing}
alpha     <- 0.3
es_fitted <- rep(NA_real_, n_obs)
es_fitted[1] <- ts_vec[1]
for (i in 2:n_obs) es_fitted[i] <- alpha*ts_vec[i-1] + (1-alpha)*es_fitted[i-1]
es_err  <- ts_vec - es_fitted

es_bias <- mean(es_err, na.rm = TRUE)
es_mad  <- mean(abs(es_err), na.rm = TRUE)
es_mse  <- mean(es_err^2,    na.rm = TRUE)
es_mape <- mean(abs(es_err / ts_vec) * 100, na.rm = TRUE)
es_rsfe <- sum(es_err, na.rm = TRUE)
es_ts   <- es_rsfe / es_mad
es_next <- alpha*ts_vec[n_obs] + (1-alpha)*es_fitted[n_obs]

cat("ES(0.3) — Bias:", round(es_bias,4), "| MAD:", round(es_mad,4),
    "| MSE:", round(es_mse,4), "| MAPE:", round(es_mape,4),
    "| RSFE:", round(es_rsfe,4), "| TS:", round(es_ts,4),
    "| Forecast:", round(es_next,4), "\n")

p_es <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,     color = "Actual")) +
  geom_line(aes(y = es_fitted, color = "ES"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","ES"="darkgreen")) +
  labs(title = "Actual vs Exponential Smoothing (α=0.3)", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_es)
ggsave("outputs/figures/exponential_smoothing_plot.png", p_es, width = 10, height = 5)
```

## 5.5 Trend-Adjusted Exponential Smoothing

```{r taes}
alpha_t <- 0.3 ; beta_t <- 0.1
lev  <- rep(NA_real_, n_obs)
trnd <- rep(NA_real_, n_obs)
lev[1] <- ts_vec[1] ; trnd[1] <- 0
for (i in 2:n_obs) {
  lev[i]  <- alpha_t*ts_vec[i]       + (1-alpha_t)*(lev[i-1]+trnd[i-1])
  trnd[i] <- beta_t*(lev[i]-lev[i-1]) + (1-beta_t)*trnd[i-1]
}
taes_fitted <- lev + trnd
taes_err    <- ts_vec - taes_fitted

taes_bias <- mean(taes_err, na.rm = TRUE)
taes_mad  <- mean(abs(taes_err), na.rm = TRUE)
taes_mse  <- mean(taes_err^2,    na.rm = TRUE)
taes_mape <- mean(abs(taes_err / ts_vec) * 100, na.rm = TRUE)
taes_rsfe <- sum(taes_err, na.rm = TRUE)
taes_ts   <- taes_rsfe / taes_mad
taes_next <- lev[n_obs] + trnd[n_obs]

cat("TAES — Bias:", round(taes_bias,4), "| MAD:", round(taes_mad,4),
    "| MSE:", round(taes_mse,4), "| MAPE:", round(taes_mape,4),
    "| RSFE:", round(taes_rsfe,4), "| TS:", round(taes_ts,4),
    "| Forecast:", round(taes_next,4), "\n")

p_taes <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,       color = "Actual")) +
  geom_line(aes(y = taes_fitted, color = "TAES"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","TAES"="firebrick")) +
  labs(title = "Actual vs Trend-Adjusted ES (α=0.3, β=0.1)",
       x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_taes)
ggsave("outputs/figures/trend_adjusted_smoothing_plot.png", p_taes, width = 10, height = 5)
```

## 5.6 Linear Trend Projection

```{r linear-trend}
t_var     <- seq_len(n_obs)
lm_fit    <- lm(ts_vec ~ t_var)
lm_fitted <- fitted(lm_fit)
lm_next   <- as.numeric(predict(lm_fit, newdata = data.frame(t_var = n_obs + 1L)))
lm_err    <- ts_vec - lm_fitted

lm_bias <- mean(lm_err)
lm_mad  <- mean(abs(lm_err))
lm_mse  <- mean(lm_err^2)
lm_mape <- mean(abs(lm_err / ts_vec) * 100)
lm_rsfe <- sum(lm_err)
lm_ts   <- lm_rsfe / lm_mad

cat("LM — Intercept:", round(coef(lm_fit)[1],4),
    "| Slope:", round(coef(lm_fit)[2],4), "\n")
cat("LM — Bias:", round(lm_bias,4), "| MAD:", round(lm_mad,4),
    "| MSE:", round(lm_mse,4), "| MAPE:", round(lm_mape,4),
    "| RSFE:", round(lm_rsfe,4), "| TS:", round(lm_ts,4),
    "| Forecast:", round(lm_next,4), "\n")

p_lm <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,     color = "Actual")) +
  geom_line(aes(y = lm_fitted, color = "Linear Trend"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","Linear Trend"="darkred")) +
  labs(title = "Actual vs Linear Trend Projection", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_lm)
ggsave("outputs/figures/trend_projection_plot.png", p_lm, width = 10, height = 5)
```

## 5.7 Seasonal Indices

```{r seasonal-indices}
# Ratio-to-moving-average seasonal indices (12 months)
n_full    <- floor(n_obs / 12) * 12          # 156
ts_mat    <- matrix(ts_vec[seq_len(n_full)], nrow = 12)
mo_means  <- rowMeans(ts_mat, na.rm = TRUE)
si        <- mo_means / mean(mo_means)       # length 12
si_rep    <- rep(si, length.out = n_obs)
deseas    <- ts_vec / si_rep

t_si      <- seq_len(n_obs)
lm_si     <- lm(deseas ~ t_si)
si_fitted <- fitted(lm_si) * si_rep
si_nt     <- as.numeric(predict(lm_si, newdata = data.frame(t_si = n_obs + 1L)))
si_next   <- si_nt * si[4]                  # April = index 4

si_err  <- ts_vec - si_fitted
si_bias <- mean(si_err, na.rm = TRUE)
si_mad  <- mean(abs(si_err), na.rm = TRUE)
si_mse  <- mean(si_err^2,    na.rm = TRUE)
si_mape <- mean(abs(si_err / ts_vec) * 100, na.rm = TRUE)
si_rsfe <- sum(si_err, na.rm = TRUE)
si_ts   <- si_rsfe / si_mad

mo_lab <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
cat("Seasonal Indices:\n")
for (i in 1:12) cat(mo_lab[i], ":", round(si[i], 4), "\n")
cat("High season:", mo_lab[which.max(si)], "\n")
cat("Low  season:", mo_lab[which.min(si)], "\n")
cat("SI — Bias:", round(si_bias,4), "| MAD:", round(si_mad,4),
    "| MSE:", round(si_mse,4), "| MAPE:", round(si_mape,4),
    "| RSFE:", round(si_rsfe,4), "| TS:", round(si_ts,4),
    "| Forecast:", round(si_next,4), "\n")

p_si <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,     color = "Actual")) +
  geom_line(aes(y = si_fitted, color = "Seasonal Indices"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","Seasonal Indices"="darkviolet")) +
  labs(title = "Actual vs Seasonal Indices Forecast", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_si)
ggsave("outputs/figures/seasonal_indices_plot.png", p_si, width = 10, height = 5)
```

## 5.8 Additive Decomposition

```{r additive-decomp}
decomp_add  <- decompose(ts_data, type = "additive")
add_fitted  <- ts_vec - as.numeric(decomp_add$random)

add_err  <- ts_vec - add_fitted
add_bias <- mean(add_err, na.rm = TRUE)
add_mad  <- mean(abs(add_err), na.rm = TRUE)
add_mse  <- mean(add_err^2,    na.rm = TRUE)
add_mape <- mean(abs(add_err / ts_vec) * 100, na.rm = TRUE)
add_rsfe <- sum(add_err, na.rm = TRUE)
add_ts   <- add_rsfe / add_mad

# Forecast trend via OLS on non-NA trend values
tr_add   <- as.numeric(decomp_add$trend)
idx_add  <- which(!is.na(tr_add))
t_idx    <- seq_along(tr_add)
lm_add   <- lm(tr_add[idx_add] ~ I(t_idx[idx_add]))
add_nt   <- coef(lm_add)[1] + coef(lm_add)[2] * (n_obs + 1)
add_next <- as.numeric(add_nt) + as.numeric(decomp_add$seasonal)[4]

cat("ADD — Bias:", round(add_bias,4), "| MAD:", round(add_mad,4),
    "| MSE:", round(add_mse,4), "| MAPE:", round(add_mape,4),
    "| RSFE:", round(add_rsfe,4), "| TS:", round(add_ts,4),
    "| Forecast:", round(add_next,4), "\n")

p_add <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,      color = "Actual")) +
  geom_line(aes(y = add_fitted, color = "Additive"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","Additive"="tomato")) +
  labs(title = "Actual vs Additive Decomposition", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_add)
ggsave("outputs/figures/additive_decomposition_plot.png", p_add, width = 10, height = 5)
```

## 5.9 Multiplicative Decomposition

```{r mult-decomp}
decomp_mult  <- decompose(ts_data, type = "multiplicative")
mult_raw     <- ts_vec / as.numeric(decomp_mult$random)
mult_fitted  <- ifelse(is.finite(mult_raw), mult_raw, NA_real_)

mult_err  <- ts_vec - mult_fitted
mult_bias <- mean(mult_err, na.rm = TRUE)
mult_mad  <- mean(abs(mult_err), na.rm = TRUE)
mult_mse  <- mean(mult_err^2,    na.rm = TRUE)
mult_mape <- mean(abs(mult_err / ts_vec) * 100, na.rm = TRUE)
mult_rsfe <- sum(mult_err, na.rm = TRUE)
mult_ts   <- mult_rsfe / mult_mad

tr_mult  <- as.numeric(decomp_mult$trend)
idx_mult <- which(!is.na(tr_mult))
t_idx    <- seq_along(tr_mult)
lm_mult  <- lm(tr_mult[idx_mult] ~ I(t_idx[idx_mult]))
mult_nt  <- coef(lm_mult)[1] + coef(lm_mult)[2] * (n_obs + 1)
mult_next <- as.numeric(mult_nt) * as.numeric(decomp_mult$seasonal)[4]

cat("MULT — Bias:", round(mult_bias,4), "| MAD:", round(mult_mad,4),
    "| MSE:", round(mult_mse,4), "| MAPE:", round(mult_mape,4),
    "| RSFE:", round(mult_rsfe,4), "| TS:", round(mult_ts,4),
    "| Forecast:", round(mult_next,4), "\n")

p_mult <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,       color = "Actual")) +
  geom_line(aes(y = mult_fitted, color = "Multiplicative"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","Multiplicative"="darkgoldenrod")) +
  labs(title = "Actual vs Multiplicative Decomposition", x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_mult)
ggsave("outputs/figures/multiplicative_decomposition_plot.png", p_mult, width = 10, height = 5)
```

## 5.10 Regression with Trend and Seasonal Dummies

```{r reg-seasonal}
month_num <- rep(1:12, length.out = n_obs)
reg_df    <- data.frame(y = ts_vec, t_var = t_var)
for (m in 2:12) reg_df[[paste0("d", m)]] <- as.integer(month_num == m)

reg_fit    <- lm(y ~ ., data = reg_df)
reg_fitted <- fitted(reg_fit)

new_reg <- data.frame(t_var = n_obs + 1L)
for (m in 2:12) new_reg[[paste0("d", m)]] <- as.integer(m == 4L)
reg_next <- as.numeric(predict(reg_fit, newdata = new_reg))

reg_err  <- ts_vec - reg_fitted
reg_bias <- mean(reg_err)
reg_mad  <- mean(abs(reg_err))
reg_mse  <- mean(reg_err^2)
reg_mape <- mean(abs(reg_err / ts_vec) * 100)
reg_rsfe <- sum(reg_err)
reg_ts   <- reg_rsfe / reg_mad

cat("REG — Bias:", round(reg_bias,4), "| MAD:", round(reg_mad,4),
    "| MSE:", round(reg_mse,4), "| MAPE:", round(reg_mape,4),
    "| RSFE:", round(reg_rsfe,4), "| TS:", round(reg_ts,4),
    "| Forecast:", round(reg_next,4), "\n")

p_reg <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,      color = "Actual")) +
  geom_line(aes(y = reg_fitted, color = "Regression"), linetype = "dashed") +
  scale_color_manual(values = c("Actual"="steelblue","Regression"="darkgreen")) +
  labs(title = "Actual vs Regression with Seasonal Dummies",
       x = "Date", y = "Index", color = "") +
  theme_minimal()
print(p_reg)
ggsave("outputs/figures/regression_seasonal_dummy_plot.png", p_reg, width = 10, height = 5)
```

# 6. Accuracy Comparison

```{r accuracy-table}
library(knitr)

accuracy_df <- data.frame(
  Method = c("Naive","Moving Average","Weighted MA","Exp Smoothing",
             "Trend-Adj ES","Linear Trend","Seasonal Indices",
             "Additive Decomp","Multiplicative Decomp","Regression+Seasonal"),
  Bias          = round(c(naive_bias, ma_bias, wma_bias, es_bias, taes_bias,
                           lm_bias,   si_bias, add_bias, mult_bias, reg_bias), 4),
  MAD           = round(c(naive_mad,  ma_mad,  wma_mad,  es_mad,  taes_mad,
                           lm_mad,    si_mad,  add_mad,  mult_mad,  reg_mad), 4),
  MSE           = round(c(naive_mse,  ma_mse,  wma_mse,  es_mse,  taes_mse,
                           lm_mse,    si_mse,  add_mse,  mult_mse,  reg_mse), 4),
  MAPE          = round(c(naive_mape, ma_mape, wma_mape, es_mape, taes_mape,
                           lm_mape,   si_mape, add_mape, mult_mape, reg_mape), 4),
  RSFE          = round(c(naive_rsfe, ma_rsfe, wma_rsfe, es_rsfe, taes_rsfe,
                           lm_rsfe,   si_rsfe, add_rsfe, mult_rsfe, reg_rsfe), 4),
  Tracking_Signal = round(c(naive_ts, ma_ts, wma_ts, es_ts, taes_ts,
                              lm_ts,  si_ts, add_ts, mult_ts, reg_ts), 4),
  Next_Forecast = round(c(naive_next, ma_next, wma_next, es_next, taes_next,
                           lm_next,   si_next, add_next, mult_next, reg_next), 4),
  stringsAsFactors = FALSE
)

kable(accuracy_df, caption = "Forecast Accuracy Comparison — All Methods")
write.csv(accuracy_df, "outputs/tables/accuracy_comparison.csv", row.names = FALSE)
cat("Accuracy table saved.\n")
```

# 7. Superior Method & Final Forecast

```{r final}
best_idx  <- which.min(accuracy_df$MAPE)
best_name <- accuracy_df$Method[best_idx]
best_fore <- accuracy_df$Next_Forecast[best_idx]

cat("=== SUPERIOR METHOD ===\n")
cat("Method    :", best_name, "\n")
cat("MAPE      :", accuracy_df$MAPE[best_idx], "%\n")
cat("MAD       :", accuracy_df$MAD[best_idx], "\n")
cat("Forecast  : April 2026 =", round(best_fore, 2), "\n")

# Match fitted values to superior method
best_fitted <- switch(best_name,
  "Regression+Seasonal"   = reg_fitted,
  "Additive Decomp"       = add_fitted,
  "Multiplicative Decomp" = mult_fitted,
  "Seasonal Indices"      = si_fitted,
  "Linear Trend"          = lm_fitted,
  "Trend-Adj ES"          = taes_fitted,
  "Exp Smoothing"         = es_fitted,
  "Weighted MA"           = wma_fitted,
  "Moving Average"        = ma_fitted,
  naive_fitted)

next_pt <- data.frame(Date = as.Date("2026-04-01"), Value = best_fore)

p_best <- ggplot(df_long, aes(x = Date)) +
  geom_line(aes(y = Value,       color = "Actual")) +
  geom_line(aes(y = best_fitted, color = "Superior Method"), linetype = "dashed") +
  geom_point(data = next_pt, aes(y = Value),
             color = "red", size = 4, shape = 17) +
  scale_color_manual(values = c("Actual"="steelblue","Superior Method"="darkgreen")) +
  labs(title    = paste("Superior Method:", best_name),
       subtitle = paste("April 2026 Forecast:", round(best_fore, 2)),
       x = "Date", y = "Index Value", color = "") +
  theme_minimal()
print(p_best)
ggsave("outputs/figures/superior_method_plot.png", p_best, width = 10, height = 5)

# Save final forecast table
final_df <- data.frame(
  Superior_Method    = best_name,
  Data_Access_Date   = "2026-05-23",
  Latest_Observation = "2026-03-01",
  Forecast_Target    = "2026-04-01",
  Forecasted_Value   = round(best_fore, 4),
  MAPE               = accuracy_df$MAPE[best_idx],
  MAD                = accuracy_df$MAD[best_idx],
  stringsAsFactors   = FALSE
)
kable(final_df, caption = "Final Forecast Result")
write.csv(final_df, "outputs/tables/final_forecast.csv", row.names = FALSE)
cat("Final forecast saved.\n")
```

# 8. Interpretation

The Export Unit Value Index reflects the average price level of Turkish exports
relative to the 2015 base year. The series displays a clear upward trend from
2020 onward driven by global commodity price increases and exchange rate
depreciation, along with mild seasonal fluctuations. The superior method
captures both trend and seasonal structure, producing a reliable one-step-ahead
forecast for April 2026.

# 9. Limitations

- TÜİK SDMX API returned HTTP 401/403 from Posit Cloud due to IP restrictions;
  the data file was accessed through the URL obtained via `tuikr`.
- Structural breaks (COVID-19 pandemic, 2021–2022 exchange rate volatility) may
  reduce model accuracy.
- External variables (exchange rates, commodity prices) are not included.
- All models assume the recent data-generating process continues into April 2026.

# 10. Reproducibility

1. Install R (≥ 4.0) and RStudio.
2. Install packages:
   ```r
   install.packages(c("devtools","readxl","httr","forecast","ggplot2","tidyr","lubridate","knitr"))
   devtools::install_github("emraher/tuikr")
   ```
3. Place `Foreign Trade Indices.xls` in the project root directory.
4. Open `forecasting_project.Rmd` and click **Knit**.
